\setchapterpreamble[u]{\margintoc}
\chapter{The Kernel}
\labch{kr}


\section{Arvo}
\labsec{kr:arvo}

We could do worse than to start our exploration of the Urbit kernel than to quote the Whitepaper itself on the subject of Arvo:

\begin{quote}
The fundamental unit of the Urbit kernel is an event called a \texttt{+\$move}.  Arvo is primarily an event dispatcher between moves created by vanes.  Each move therefore has an address and other structured information.  (\cite{Yarvin2017})
\end{quote}

Every vane defines its own structured events (\texttt{+\$move}s).  Each unique kind of structured event has a unique, frequently whimsical, name.  This can make it challenging to get used to how a particular vane behaves.

\marginnote[2mm]{The system-level instrumentation of Arvo is described in Chapter~\ref{support}.}

Arvo is essentially an event handler which can coordinate and dispatch messages between vanes as well as emit \unix~events (side effects) to the underlying (presumed Unix-compatible) host OS.  Arvo as hosted OS does not carry out any tasks specific to the machine hardware, such as memory allocation, system thread management, and hardware- or firmware-level operations.  These are left to the \emph{king} and \emph{serf}, the daemon processes which together run Arvo.

Arvo is architected as a state machine, the deterministic end result of the event log.  We need to examine Arvo from two separate angles:

\begin{enumerate}
  \item  Event processing engine and state machine (vane coordinator)
  \item  Standard noun structure (“Arvo-shaped noun”)
\end{enumerate}

\subsection{Event Processing Engine}
\labsec{kr:arvo-state}

\begin{quote}
A vanilla event loop scales poorly in complexity.  A system event is the trigger for a cascade of internal events; each event can schedule any number of future events.  This easily degenerates into “event spaghetti.”  Arvo has “structured events”; it imposes a stack discipline on event causality, much like imposing subroutine structure on \texttt{goto}s.  (\cite{Yarvin2017})
\end{quote}

Arvo events are known as \texttt{++move}s, containing metadata and data.  Arvo recognizes four types of moves:

\begin{enumerate}
  \item  \pass~events are forward calls from one vane to another (or back to itself, occasionally), and
  \item  \give~events are returned values and move back along the calling duct.
  \item  \slip~events [TODO think of like ref counting?]
  \item  \unix~events communicate from Arvo to the underlying binary in such a way as to emit an external effect (an \ames~network communication, for instance, or text input and output).
\end{enumerate}

show structure of a card

"Each vane defines a protocol for interacting with other vanes (via Arvo) by defining four types of cards: tasks, gifts, notes, and signs."
"In other words, there are only four ways of seeing a move: (1) as a request seen by the caller, which is a note. (2) that same request as seen by the callee, a task. (3) the response to that first request as seen by the callee, a gift. (4) the response to the first request as seen by the caller, a sign."
(TODO move trace work here)

Without reference to the particular content of vanes, let us briefly diagram a “move trace”, or examination of how an event generated by a vane produces results via Arvo.

TODO remote call via \ames?

\begin{quote}
"An interrupted event never happened.  The computer is deterministic; an event is a transaction; the event log is a log of successful transactions. In a sense, replaying this log is not Turing complete. The log is an existence proof that every event within it terminates.  (\cite{Yarvin2017})
\end{quote}

\subsection{Standard Noun Structure}
\labsec{kr:arvo-noun}

Arvo defines five standard arms for vanes and the binary runtime to use:

\begin{enumerate}
  \item  \texttt{++peek} grants read-only access to \clay; this is called a \emph{scry}.
  \item  \texttt{++poke} accepts \texttt{++move}s and processes them; this is the only arm that actually alters Arvo's state.
  \item  \texttt{++wish} accepts a core and parses it against \zuse.
  \item  \texttt{++come} and \texttt{++load} are used in kernel upgrades, allowing Arvo to update itself in-place.
\end{enumerate}


\href{Arvo tutorial}{https://urbit.org/docs/arvo/overview/}

\subsection[\zuse~\&~\lull]{\zuse~and \lull}
\labsec{zuse}

\zuse~and \lull~define common structures and library functions for Arvo.

subject wrapped

\section{Arvo Vanes}
\labsec{kr:vanes}

Each vane has a characteristic structure which identifies it as a vane to Arvo and allows it to handle \move s consistently.

\section[\ames]{\ames, A Network}
\labsec{kr:ames}

In a sense, \ames~is the operative definition of an urbit on the network.  That is, from outside of one's own urbit, the only specification that must be hewed to is that \ames~behaves a certain way in response to events.

\ames~implements a system expecting—and delivering—guaranteed one-time delivery.  This derives from an observation by \citeauthor{Yarvin2016}~in the Whitepaper:  "bus v. commands whatever"

UDP packet structure

network events
acks \& nacks

\subsection{Scrying into \ames}
\labsec{kr:a:scry}

\ames~scry

\begin{table}[h!]
  \begin{center}
    \caption{\ames~\dotket~Calls.}
    \label{ha:ames}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%x} & Get ship and peer information:  protocol version, peers, ship state, etc. & \\
    \end{tabular}
  \end{center}
\end{table}


\section[\behn]{\behn, A Timer}
\labsec{behn}

\behn~is a simple vane that promises to emit events after—but never before—their timestamp.  This guarantee

As the shortest vane, we commend \behn~to the student as an excellent subject for a first dive into the structure of a vane.

\behn~maintains an event handler and a state.

Any task may have one of the following states:

\begin{lstlisting}
%born  born:event-core
%rest  (rest:event-core date=p.task)
%drip  (drip:event-core move=p.task)
%huck  (huck:event-core syn.task)
%trim  trim:event-core
%vega  vega:event-core
%wait  (wait:event-core date=p.task)
%wake  (wake:event-core error=~)
\end{lstlisting}

\subsection{Scrying into \behn}
\labsec{kr:b:scry}

\begin{table}[h!]
  \begin{center}
    \caption{\behn~\dotket~Calls.}
    \label{ha:behn}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%x} & Get timers, timestamps, next timer to fire, etc. & \\
    \end{tabular}
  \end{center}
\end{table}

\section[\clay]{\clay, A File System}
\labsec{clay}

```
~tinnus-napbus
3:57 PM
what is the correct way to read a file on a remote ship? I've tried both warp and werp and I'm not getting a response, just messages in the target about clay something something indirect and then I get crash on fragment errors sometimes
with an %x care
~rovnys-ricfer
4:02 PM
warp should work, I think
make sure the file is actually there
i.e. can you do the same warp on the local ship?
```

\begin{itemize}
  \item  \texttt{aeon} is
  \item  \texttt{arch} is
  \item  \texttt{care} is the \clay~submode, defined in \lull.
  \item  \texttt{desk} is
  \item  \texttt{dome} is
  mark cast file
\end{itemize}


\subsection{Scrying into \clay}
\labsec{kr:c:scry}

\clay~has the most sophisticated scry taxonomy.

\begin{table}[h!]
  \begin{center}
    \caption{\clay~\dotket~Calls.}
    \label{ha:clay}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%a} & Expose file build into namespace. & \texttt{.\^(vase \%ca /\~zod/home/1/lib/generators/hoon} \\
      \texttt{\%b} & Expose mark build into namespace. & \\
      \texttt{\%c} & Expose cast build into namespace. & \\
      %\texttt{\%d} & & \\
      \texttt{\%e} & & \\
      \texttt{\%f} & & \\
      \texttt{\%p} & Get the permissions that apply at path. & \\
      \texttt{\%r} & Get the data at a node (like \texttt{\%x}) and wrap it in a vase. & \\
      \texttt{\%s} & produce yaki or blob for given tako or lobe & \\
      \texttt{\%t} & produce the list of paths within a yaki with :pax as prefix & \\
      \texttt{\%u} & Check for existence of a node at aeon. & \\
      \texttt{\%v} & Get the desk state at a specified aeon. & \\
      \texttt{\%w} & Get all cases referring to the same revision as the given case. & \\
      \texttt{\%x} & Get the data at a node. & \\
      \texttt{\%y} & Get the \texttt{arch} (directory listing) at a node. & \\
      \texttt{\%z} & Get a recursive hash of a node and its children. & \\
    \end{tabular}
  \end{center}
\end{table}

% https://github.com/urbit/urbit/blob/master/pkg/arvo/sys/vane/gall.hoon#L1538
% https://urbit.org/blog/ford-fusion/


\subsection[\ford]{\ford, A Build System}
\labsec{ford}

runes
%https://urbit.org/blog/ford-fusion/

\begin{table}[h!]
  \begin{center}
    \caption{\ford~Runes.}
    \label{ha:ford}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{/-} & Import structure file from \texttt{sur}. & \\
      \texttt{/+} & Import library file from \texttt{lib}. & \\
      \texttt{/=} & Import user-specified file. & \\
      \texttt{/*} & Import contents of file converted by mark. & \\
    \end{tabular}
  \end{center}
\end{table}

importing with \texttt{*} is w/o face, foo=bar


\subsection[Marks]{Marks and conversions}
\labsec{marks2}

\section[\dill]{\dill, A Terminal driver}
\labsec{dill}

\subsection{Scrying into \dill}
\labsec{kr:d:scry}

\dill~scrys are unusual, in that they are typically only necessary for fine-grained Arvo control of the display.  Even command-line apps instrumented with \shoe~do not call into \dill~commonly.  The only instance of use in the current Arvo kernel is in Herm, the terminal session manager.

\begin{table}[h!]
  \begin{center}
    \caption{\dill~\dotket~Calls.}
    \label{ha:dill}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%x} & Get the current line or cursor position of default session. & \\
    \end{tabular}
  \end{center}
\end{table}


\section[\eyre~\&~\iris]{\eyre~and \iris, Server and Client Vanes}
\labsec{eyre}

\subsection{Scrying into \eyre}
\labsec{kr:e:scry}

\eyre~

\begin{table}[h!]
  \begin{center}
    \caption{\jael~\dotket~Calls.}
    \label{ha:iris}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%x} & Get CORS etc TODO. & \\
      \texttt{\%\$} & Get . & \\  % TODO is this default subject?
    \end{tabular}
  \end{center}
\end{table}

\subsection{Scrying into \iris}
\labsec{kr:i:scry}

\begin{table}[h!]
  \begin{center}
    \caption{\iris~\dotket~Calls.}
    \label{ha:iris}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%\$} & Get . & \\  % TODO is this default subject?
    \end{tabular}
  \end{center}
\end{table}

\section[\jael]{\jael, Secretkeeper}
\labsec{jael}

\marginnote[2mm]{
\jael is named after Jael, the wife of Heber, who kept mum and slew fleeing enemy general Sisera in Judges 4.
}

\jael~keeps secrets, the cryptographic keys that make it possible to securely control your Urbit.

\jael~weighs in as one of the shorter vanes

\begin{lstlisting}

\end{lstlisting}


\subsection{Scrying into \jael}
\labsec{kr:j:scry}




\marginnote[2mm]{
Sometimes residual elements of vane development leak into release code and yield insight into how the kernel developers produce new vanes.  For instance, when a new version of \jael~was being developed, it was dubbed \texttt{kale} and used the scry path \texttt{\%k}.  This made it into a release version of \texttt{lib/ring} in Arvo 309 K.
}

\begin{table}[h!]
  \begin{center}
    \caption{\jael~\dotket~Calls.}
    \label{ha:jael}
    \begin{tabular}{lll}
      Symbol & Meaning & Example \\
      \hline \\
      \texttt{\%\$} & Get TODO. & \\
      \texttt{\%\$} & Get . & \\
    \end{tabular}
  \end{center}
\end{table}




\section[Azimuth]{Azimuth, Address Space Management}
\labsec{azimuth2}

Urbit HD wallet


\paragraph{Comet keys}.  Comets do not have an associated Urbit HD wallet, and their keys work slightly differently.

> yeah thats how comet mining works. so you'd just put the private key you generated for a comet on the card, and this would be the ames DH exchange private key. i suppose you could still obfuscate it with a master ticket @q, by just picking a 128 bit hash, but it would be used differently than a normal azimuth master key, which is a @q used to derive a bunch of ethereum wallets private keys (and ultimately the initial network key, but that isnt required).
9:15
>and yeah whether a key works at the time it is mined is dependent on whether the routing node automatically assigned to the comet public key is currently working
>~lagrev-nocfep: the comets name is its public key


\section[Hoon Parser]{The Hoon Parser}
\labsec{hoonparse}
